The Constant-Q Transform Spectral Envelope Coefficients: A Timbre Feature Designed for Music

### 2021-07-06 IEEE SP Magazine/07-2021 - Draft/article.tex
Working on the computational example section.

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be practical in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily well adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 85 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but typically in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize timbre by MIR practitioners.

@@

We propose the CQT-SECs, a novel timbre feature which is well adapted to musical data, pitch-independent, simple to compute, interpretable, and functional. We will show how to derive it from the CQT, a frequency transform with a logarithmic resolution which matches the notes of the equal temperament, a musical scale typically used in Western music \cite{brown1991, brown1992}. We will first show how to decompose the CQT spectrum into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extract a number of timbral coefficients from the spectral envelope.

@@

Figure \ref{fig:extraction} shows an example of CQT-SECs, on the left plot. 20 coefficients were extracted from the envelope resulting from the deconvolution of the CQT spectrogram of the musical signal shown in Figure \ref{fig:deconvolution}. These coefficients essentially correspond to the harmonics of the musical instrument which contain most of its spectral energy and can therefore be a good representation of its timbre. For comparison, we also show the MFCCs computed from the same musical signal, on the right plot. 20 coefficients were computed using librosa \cite{mcfee2015}, with a sampling rate of 16 kHz, a window length of 1024 samples, and hop length of 512 samples.

We recall below how the MFCCs are commonly derived \cite{mermelstein1976}. First, a classic spectrogram is computed from the audio signal using the short-time Fourier transform (STFT) and the powers of the frequencies in each frame are then mapped onto the mel scale \cite{stevens1937} using triangular overlapping windows. This essentially produces a mel spectrogram. Then, the logs of all the mel frequencies are taken, followed by the discrete cosine transform (DCT) of each frame. Finally, the MFCCs are selected as the lowest coefficients in the resulting spectrum, excluding the very first one (DC component). This process is supposed to decouple the envelope from the pitch in the time domain, by extracting the slow-varying time components, which most likely correspond to the envelope and which will become the lowest coefficients, from the fast-varying time components, which most likely correspond to the pitch and which will become the higher coefficients.

@@

We propose to measure the discriminative power of the CQT-SECs by evaluating them on the NSynth dataset, a large-scale of annotated musical notes which is publicly available \cite{engel2017}.

Explain NSynth
Compare to the MFCCs
CQT-SECs computed with some parameters
MFCCs with some parameters

>>>>

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be practical in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily well adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 85 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but typically in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, \cite{hung2019}, \cite{luo2019}, \cite{lee2020} and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize timbre by MIR practitioners.

@@

We present the CQT-SECs, a novel timbre feature which is well adapted to musical data, pitch-independent, simple to compute, interpretable, and functional. We will show how to derive it from the CQT, a frequency transform with a logarithmic resolution which matches the notes of the equal temperament, a musical scale typically used in Western music \cite{brown1991, brown1992}. We will first show how to decompose the CQT spectrum into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extract a number of timbral coefficients from the spectral envelope.

@@

Figure \ref{fig:extraction} shows an example of CQT-SECs, on the left plot. 20 coefficients were extracted from the envelope resulting from the deconvolution of the CQT spectrogram of the musical signal shown in Figure \ref{fig:deconvolution}. These coefficients essentially correspond to the harmonics of the musical instrument which contain most of its spectral energy and can therefore be a reasonable representation of its timbre. For comparison, we also show the MFCCs computed from the same musical signal, on the right plot. 20 coefficients were computed using librosa \cite{mcfee2015}, with a sampling rate of 16 kHz, a window length of 1024 samples, and hop length of 512 samples, matching the time resolution of the CQT-SECs so that both features have the same size in time and frequency.

We recall below how the MFCCs are commonly derived. First, a classic spectrogram is computed from the audio signal using the short-time Fourier transform (STFT) and the powers of the frequencies in each frame are then mapped onto the mel scale \cite{stevens1937} using triangular overlapping windows. This essentially produces a mel spectrogram. Then, the logs of all the mel frequencies are taken, followed by the discrete cosine transform (DCT) of each frame. Finally, the MFCCs are selected as the lowest coefficients in the resulting spectrum, excluding the very first one (DC component). This process is supposed to decouple the envelope from the pitch in the time domain, by extracting the slow-varying time components, which most likely correspond to the envelope and which will become the lowest coefficients, from the fast-varying time components, which most likely correspond to the pitch and which will become the highest coefficients \cite{mermelstein1976}.

@@

We propose to measure the discriminative power of the CQT-SECs by evaluating them on the NSynth dataset, a large-scale and high-quality dataset of annotated musical notes which is publicly available\footnote{https://magenta.tensorflow.org/datasets/nsynth} \cite{engel2017}. The NSynth dataset is composed of 305,979 musical notes which were generated from 1,006 instruments as 4-second, monophonic audio signals at a sampling rate of 16 kHz, with pitches ranging over all the note numbers of a standard MIDI piano (21-108) and with 5 different velocities (25, 50, 75, 100, and 127), whenever applicable. The instruments are organized into 11 families (bass, brass, flute, guitar, keyboard, mallet, organ, reed, string, synth\_lead, and vocal) and 3 sources (acoustic, electronic, and synthetic). We chose to evaluate the CQT-SECs on the notes with a velocity of 75 only, leading to 60,388 different notes for 945 different instruments.

We derived the CQT-SECs for all the notes in this subset, by first computing the CQT spectrogram using librosa, with a sampling rate of 16 kHz, a hop length of 512 samples, a minimum frequency of 32.70 Hz (corresponding to C1), 95 frequency bins, and 12 bins per octave, and then extracting 20 coefficients from the envelope resulting from the deconvolution of the CQT spectrogram. We used a power CQT spectrogram (i.e., magnitude to the power of 2) and took the real part of the envelope to ensure real values. For comparison, we also computed the MFCCs using librosa, with a sampling rate of 16 kHz, a window length of 1024 samples, a hop length of 512 samples, and 20 coefficients, matching the time resolution of the CQT-SECs so that both features have the same size in time and frequency.
